---
title: "Bayesian Modelling of Urban Observatory Data"
author: "Jonny Law"
date: "10 January 2017"
output: 
  revealjs::revealjs_presentation:
    theme: moon
    controls: FALSE
    progress: FALSE
    css: my_presentation.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(latex2exp)
library(ggthemes)
theme_set(theme_solarized_2(light = FALSE))
```

## Environment Sensors

```{r read_data, cache=TRUE}
emote = read_csv("data/sensor_data.csv")
```


```{r plot_environment_sensor}
emote %>%
  mutate(Variable = paste0(Variable, " (", Units, ")")) %>%
  ggplot(aes(x = Timestamp, y = Value, colour = Variable)) +
  geom_point(size = 0.1) +
  facet_wrap(~Variable, scales = "free_y") +
  theme(legend.position = "none") +
  labs(title = "Environmental Sensor Readings", subtitle = "Raw readings of sensor data at emote_1108")
```

## A Naive Model

* Propose a state space model for each time series

$$\begin{aligned}
y_{ti} &= F_{ti}x_{ti} + v_{ti}, &v_{ti} &\sim \mathcal{N}(0, V_i), \\
x_{ti} &= G_{ti}x_{t-1} + w_{ti}, &w_{ti} &\sim \mathcal{N}(0, W_i), \\
x_0 &\sim p(x_0).
\end{aligned}$$

* Couple the models together using block diagonal matrices, for $i = 2$

$$F_t = \begin{bmatrix}F_{t1} &0 \\ 0 &F_{t2}\end{bmatrix}, G_t = \begin{bmatrix}G_{t1} & 0 \\ 0 & G_{t2}\end{bmatrix}.$$

## Learning Covariance Matrices

> - In the Naive model the system covariance of the $i^{\textrm{th}}$ model has dimension $d_i \times d_i$
> - If we model the four time series, the system covariance matrix contains $\sum_{i=1}^4 d_i^2$ terms
> - *This is too large!*

## Factor Analysis

$$y_t = \beta^T f_t + \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0, \sigma_t)$$

> - $y_t$ is a $p$-dimensional
> - $\beta$ is a $k \times p$ matrix, $k << p$
> - $f_t \sim \mathcal{N}(0, I_k)$ are latent variables
> - $\operatorname{Var}(y_t) = \beta^\prime\beta + \sigma_t$

----

```{r sim_fa}
set.seed(1)
n = 100
p = 6
k = 2
sigma = diag(runif(p))
y = matrix(NA_real_, nrow = n, ncol = p)
f = MASS::mvrnorm(n, mu = rep(0, k), Sigma = diag(k))
eps = MASS::mvrnorm(n, mu = rep(0, p), Sigma = sigma)
beta = matrix(runif(p * k), nrow = p, ncol = k)
beta[1,1] = 1
beta[2,2] = 1
beta[1,2] = 0
for (i in 1:n) {
  y[i,] = beta %*% f[i,] + eps[i,] 
}
factor_sims = as.data.frame(y)
factor_sims = cbind(factor_sims, f)
colnames(factor_sims) = c(paste0("y", 1:6), paste0("f", 1:2))
```

<div id = "left">
$$\beta^T = \begin{pmatrix}
1 & 0 \\
0.31 & 1 \\
0.27 & 0.84 \\
0.13 & 0.88 \\
0.43 & 0.77 \\
0.70 & 0.32
\end{pmatrix}$$
</div>

<div id = "right">
```{r plot_fa}
factor_sims %>%
  add_column(time = 1:nrow(factor_sims)) %>%
  gather(key, value, contains("y")) %>%
  ggplot(aes(x = time, y = value, colour = key)) +
  geom_line() +
  facet_wrap(~key) +
  theme(legend.position = "none") +
  labs(title = "Simulation from Factor Model")
```
</div>

## Time Varying Variance: Stochastic Volatility Model

$$\begin{aligned}
y_t &= \varepsilon_t \exp(\alpha_t), &\varepsilon_t &\sim \mathcal{N}(0, 1) \\
\alpha_t &= \phi (\mu - \alpha_t) + \eta_t, &\eta_t &\sim \mathcal{N}(0, \sigma_\eta). \\
\alpha_0 &\sim p(\alpha_0).
\end{aligned}$$

> - $y_t$ is a univariate time-series observed at times $t = t_0,\dots,t_T$
> - $\alpha_t$ is the log-volatility, modelled using an AR(1) process
> - Write this as $y_t \sim \textrm{ISV}_T(\phi; \sigma_\eta; \mu)$

----

```{r sim_ar1}
ar1 = function(n, phi, sigma, mu) {
  x = numeric(n)
  x[1] = rnorm(1)
  for (i in 2:n) {
    x[i] = x[i-1] + phi * (mu - x[i-1]) + rnorm(1) * sigma
  }
  x
}
```

```{r sim_isv}
isv = function(n, alpha) {
  y = numeric(n)

  for (i in 1:n) {
    y[i] = rnorm(1) * exp(alpha[i])
  }
  data_frame(time = 1:n, y, alpha)
} 
n = 1000
alpha = ar1(n, 0.05, 0.4, 0.0)
isv_sims = isv(n, alpha)
```

```{r plot_isv}
isv_sims %>%
  gather(key, value, -time) %>%
  ggplot(aes(x = time, y = value, colour = key)) +
  geom_line() +
  facet_wrap(~key, ncol = 1, scales = "free") +
  theme(legend.position = "none") +
  labs(title = "Simulation from a Stochastic Volatility Model", 
       subtitle = TeX("$\\sigma = 1.0, \\phi = 0.05, , \\sigma_\\eta = 0.4$"))
```

## Factor Stochastic Volatility

$$\begin{aligned}
y_t &= \beta f_t + v_t, \quad v_t \sim \mathcal{N}(0, V), \\
f_t &\sim \textrm{ISV}(\phi; \sigma_\eta; \mu).
\end{aligned}$$

> - Combine factor analysis and stochastic volatility models to learn about many related processes with non-constant variance
> - This model has fewer parameters than the naive model
> - Parameter learning is performed using Gibbs Sampling and Kalman Filtering

----

```{r factor_sv}
fsv = function(n, v, beta, ft) {
  k = ncol(beta)
  p = nrow(beta)
  y = matrix(NA_real_, ncol = p, nrow = n)
  for (i in 1:n) {
    y[i,] = beta %*% ft[i,] + rnorm(1, sd = sqrt(v))
  }
  out = cbind(1:n, y)
}
```

```{r sim_fsv}
ft = replicate(k, {
  alpha = ar1(n, 0.05, 0.4, 0.0)
  isv(n, alpha)$y
  })
fsv_sims = fsv(n, 0.1, beta, ft) %>% as.data.frame()
colnames(fsv_sims) = c("time", paste0("y", 1:6))
```

```{r plot_fsv}
fsv_sims %>%
  gather(key, value, -time) %>%
  ggplot(aes(x = time, y = value, colour = key)) +
  geom_line() +
  facet_wrap(~key) +
  theme(legend.position = "none") +
  labs(title = "Simulation from the Factor Stochastic Volatility Model",
       subtitle = TeX("$\\sigma = 1.0, \\phi = 0.05, , \\sigma_\\eta = 0.4$"))
```

## Non Centered Factor SV

$$\begin{aligned}
y_t &= F_{t}\mu_{t} + \beta f_t + \varepsilon_t, &\varepsilon_t &\sim \mathcal{N}(0, \sigma) \\
x_t &= G_tx_{t-1} + w_{t}, &w_t &\sim \mathcal{N}(0, W) \\
f_t &\sim \textrm{ISV}(\phi; \sigma_\eta; \mu).
\end{aligned}$$

> - A DLM is used to model the mean behaviour of each time series
> - The matrices $F_t$ and $G_t$ are block diagonal
> - $W = \operatorname{diag}(\sigma_{w,1}, \dots, \sigma_{w, pd})$

----

## Thanks

* Use my Scala package: Bayesian inference for D(G)LMs [git.io/dlm](https://git.io/dlm)
* Read our paper "Composable models for online Bayesian analysis of streaming data" published in Statistics and Computing (DOI 10.1007/s11222-017-9783-1.)